Toronto Paramedic Services

### 2024-09-26 paper/paper.qmd
Edited the data section to include explanations of shapefiles, added an appendix, and added explanations of implications of results.

As the objective is to understand the driving factors of EMS demand, the actual data that is used within this analysis reflects the total EMS demand and the demand for each incident type over the course of six years (2017-2022). All recorded incidents were aggregated together by incident type to see the number of emergency calls for each type. Then, the number of emergency calls for each incident type were summed to produce a total EMS demand count. This process was repeated every year from 2017 to 2022. @tbl-analysisdata (found below) illustrates the outcome of this process, and summary statistics for this analysis data can be found in the Appendix (@sec-appendix).

```{r}
#| echo: false
#| message: false
#| label: tbl-analysisdata
#| tbl-cap: "Organized Paramedic Services Incident Data (2017-2022)"

@@

data |>
  slice(1:6) |>
  kable(
    col.names = c("Year", "Medical Emergency Calls", "Vehicle Accident Calls", "Emergency Transfer Calls",
                  "Fire Calls", "Airport Standby Calls", "Unclassified Calls", "Total Emergency Calls"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "l", "l", "l", "l")
  )
```

The column "Unclassified Calls" in @tbl-analysisdata represents the emergency calls that were not assigned an incident type. It is important to note that throughout the analysis, these records are omitted.

@@

To further understand what drives EMS demand, this analysis uses the Forward Sortation Areas (FSA) provided by the Paramedic Services Incident Data (2017-2022) to map the driver of EMS demand within each FSA. For the maps within this analysis, the analysis data explained in @sec-analysis needed to be modified. By grouping the data by FSAs, the types of incidents recorded within each unique FSA and the number of incidents that occurred with each incident type could be found. The "driver" of EMS demand for a given FSA was determined by the incident type that had the most number of calls. A small snippet of the modified analysis data is shown below in @tbl-mapdata.

@@

To map this data, a shapefile with the boundaries for Toronto's FSAs was required. This data was acquired from the Open Government Data portal by the Government of Canada (CITATION).

@@

ggplot(data, aes(x = Year, y = Total_Emergency_Calls)) + 
  geom_line(linewidth = 0.5, colour="black") + geom_point(colour = "black") +
  ylim(200000,300000) + theme_light() + 
  ggtitle("EMS Demand From 2017 to 2022") +
  labs(x = "Year", y = "Demand for EMS", caption = "Data Source: Toronto Paramedic Services") + 
  theme(plot.title = element_text(hjust = 0.5)) 
```
@fig-totalemsdemand displays the total number of emergency calls for each year. The total number of emergency calls appears to increase in 2020 after a notable decrease in 2019. This pattern likely reflects the the onset of the covid-19 pandemic. Though the number of emergency calls appears to fluctuate over the years, the change is not drastic, indicating that there has consistently been a heightened demand for EMS within Toronto.

@@

To further unpack what is driving this consistent demand, the calls for individual incident type are also considered. @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls. Based on @fig-incidentemsdemand, demands for EMS are markedly higher for medical emergencies than any other incident type, suggesting that medical circumstances are the key drivers that propel EMS demand.

By zooming into @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided. With @fig-incidentemsdemandzoom, it becomes clear that a second prominent driver of EMS demand is vehicle accidents, implying that a fairly large portion of the emergency calls that are dispatched to paramedic services comes from incidents on the road.

@@

data |>
  pivot_longer(
    cols = c("Total_Emergency_Calls", "Medical_Emergency_Calls", "Vehicle_Accident_Calls", "Emergency_Transfer_Calls", "Fire_Calls", "Airport_Standby_Calls"),
    names_to = "series",
    values_to = "value"
  ) |>
  ggplot(mapping = aes(x = Year, y = value, colour = series)) +
  geom_line() +
  geom_point() +
  theme_light() +
  labs(
    x = "Year", colour = "Incident Type",
    caption = "Data Source: Toronto Paramedic Services"
  ) +
  scale_colour_brewer(palette = "Dark2", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Medical Emergencies", "Total Emergency Calls", "Vehicle Accidents")) +
  ggtitle("EMS Demand (2017-2022) Based on Incident Type") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylim(0,300000) + scale_y_continuous(name = "Demand for EMS", labels = scales::comma)
```
```{r}
#| echo: false 
#| message: false
#| label: fig-incidentemsdemandzoom
#| fig-cap: EMS Demand for Each Year between 2017 and 2022

@@

@@

@@

toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/TorontoForwardSortationAreas")
#view(toronto_forward_sortation_areas)

@@

As demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help better inform strategies to better manage it. To this, the given analysis finds that medical emergencies and vehicle accidents are notable drivers of EMS demand within Toronto.

Response times for EMS demand have been greatly impacted by offload delays, staffing/training challenges, and traffic congestion [@auditorreport]. With medical emergencies, a shortage of doctors has placed pressure on paramedic services as residents rely on and turn to these services more often [@cbcarticle]. This in turn, increases in offload delays - the time it takes to offload patients into hospitals. Longer offload delays mean longer response times for other incidents. Putting all of this together, a strategy to better manage EMS demand for medical emergencies may be to focus on promoting primary-care providers (i.e. family doctors) and to allocate resources to provide additional health services (e.g. programs to consult with a health-care professional online from home).

Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected. To efficiently navigate EMS demand for vehicle accidents, strategies may center around accident prevention or effective distribution of resources near accident-prone regions for faster response times.

@@

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-analysisdata
#| tbl-cap: "Organized Paramedic Services Incident Data (2017-2022)"

@@

summary_data <- summary(data[1,2:8])
view(summary_data)

data |>
  slice(1:6) |>
  kable(
    col.names = c("Year", "Medical Emergency Calls", "Vehicle Accident Calls", "Emergency Transfer Calls",
                  "Fire Calls", "Airport Standby Calls", "Unclassified Calls", "Total Emergency Calls"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "l", "l", "l", "l"),
    format.args = list(big.mark = ",")
  )

@@

@@

>>>>

As the objective is to understand the driving factors of EMS demand, the actual data that is used within this analysis reflects the total EMS demand and the demand for each incident type over the course of six years (2017-2022). All recorded incidents were aggregated together by incident type to see the number of emergency calls for each type. Then, the number of emergency calls for each incident type were summed to produce a total EMS demand count. This process was repeated every year from 2017 to 2022. @tbl-analysis (found below) illustrates the outcome of this process, and summary statistics for this analysis data can be found in the Appendix (@sec-appendix).

```{r}
#| echo: false
#| message: false
#| label: tbl-analysis
#| tbl-cap: "Organized Paramedic Services Incident Data (2017-2022)"

@@

data |>
  slice(1:6) |>
  kable(
    col.names = c("Year", "Medical Calls", "Vehicle Accident Calls", "Emergency Transfer Calls",
                  "Fire Calls", "Airport Standby Calls", "Unclassified Calls", "Total Emergency Calls"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "l", "l", "l", "l")
  )
```

The column "Unclassified Calls" in @tbl-analysis represents the emergency calls that were not assigned an incident type. It is important to note that throughout the analysis, these records are omitted.

@@

To further understand what drives EMS demand, this analysis uses the Forward Sortation Areas (FSA) provided by the Paramedic Services Incident Data (2017-2022) to map the driver of EMS demand within each FSA. For the maps within this analysis, the analysis data explained in @sec-analysis needed to be modified. By grouping the data by FSAs, the types of incidents recorded within each unique FSA and the number of incidents that occurred with each incident type could be found. The "driver" of EMS demand for a given FSA was determined by the incident type that had the most number of calls. A small snippet of the modified analysis data is shown below in @tbl-mapdata. The column that contains each FSA is named "CFSAUID" to ensure that each FSA can be mapped.

@@

To map this data, a shapefile with the boundaries for Toronto's FSAs was required. This data was acquired from the Open Government Data portal by the Government of Canada [@2021census]. Ambulance station location data was also considered in this analysis. This dataset provided by Toronto Paramedic Services [@torontoparamedicservices] and shared through the city of Toronto's Open Data portal [@opentorontodata], contains the exact geographic locations of each ambulance station within Toronto. Additional information about these datasets, including how they were prepared and used can be found in the Appendix (@sec-appendix).

@@

ggplot(data, aes(x = Year, y = Total_Emergency_Calls)) + 
  geom_line(linewidth = 0.5, colour="black") + geom_point(colour = "black") +
  ylim(200000,300000) + theme_light() + 
  ggtitle("EMS Demand From 2017 to 2022") +
  labs(x = "Year", y = "Demand for EMS", caption = "Data Source: Toronto Paramedic Services") + 
  theme(plot.title = element_text(hjust = 0.5)) 
```

@fig-totalemsdemand displays the total number of emergency calls for each year. The total number of emergency calls appears to increase in 2020 after a notable decrease in 2019. This pattern likely reflects the the onset of the covid-19 pandemic. Though the number of emergency calls appears to fluctuate over the years, the change is not drastic, indicating that there has consistently been a heightened demand for EMS within Toronto.

@@

To further unpack what is driving this consistent demand, the calls for individual incident type are also considered. @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls. Based on @fig-incidentemsdemand, demands for EMS are markedly higher for medical emergencies than any other incident type, suggesting that medical circumstances are key drivers that propel EMS demand.

@@

data |>
  pivot_longer(
    cols = c("Total_Emergency_Calls", "Medical_Emergency_Calls", "Vehicle_Accident_Calls", "Emergency_Transfer_Calls", "Fire_Calls", "Airport_Standby_Calls"),
    names_to = "series",
    values_to = "value"
  ) |>
  ggplot(mapping = aes(x = Year, y = value, colour = series)) +
  geom_line() +
  geom_point() +
  theme_light() +
  labs(
    x = "Year", colour = "Incident Type",
    caption = "Data Source: Toronto Paramedic Services"
  ) +
  scale_colour_brewer(palette = "Dark2", labels = c("Airport Standbys", "Emergency Transfers", "Fires", "Medical Emergencies", "Total Emergency Calls", "Vehicle Accidents")) +
  ggtitle("EMS Demand (2017-2022) Based on Incident Type") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylim(0,300000) + scale_y_continuous(name = "Demand for EMS", labels = scales::comma)
```

By zooming into @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided. With @fig-incidentemsdemandzoom, it becomes clear that a second driver of EMS demand may be vehicle accidents, though not as strong as medical emergencies.

```{r}
#| echo: false 
#| message: false
#| label: fig-incidentemsdemandzoom
#| fig-cap: EMS Demand for Each Year between 2017 and 2022

@@

As a prominent driver of EMS demand is medical-emergency-related calls, strategies that focus on efficiently managing calls for medical emergencies may help to better navigate EMS demand as a whole. Response times for EMS demand have been greatly impacted by offload delays, staffing/training challenges, and traffic congestion [@auditorreport]. With medical emergencies, a shortage of doctors has placed pressure on paramedic services as residents rely on and turn to these services more often [@cbcarticle]. This in turn, increases in offload delays - the time it takes to offload patients into hospitals. Longer offload delays mean longer response times for other incidents. Putting all of this together, a strategy to better manage EMS demand for medical emergencies may be to focus on promoting primary-care providers (i.e. family doctors) and to allocate resources to provide additional health services (e.g. programs to consult with a health-care professional online from home).

Similarly, as vehicle accidents are a second potential driver of EMS demand, overall EMS demand may be better managed if paramedic service strategies also target vehicle-accident-related calls. Traffic congestion has been found to increase paramedic response times [@journal], and with the current increase in traffic due to multiple construction projects throughout Toronto, longer paramedic response times and the unavailability of paramedic services can be expected. To efficiently navigate EMS demand for vehicle accidents, strategies may center around accident prevention or effective distribution of resources near accident-prone regions for faster response times.

@@

By examining EMS demand through different incident types, the investigation above finds that medical emergencies...

@@

toronto_forward_sortation_areas <- st_read("~/Toronto_Paramedic_Services/data/TorontoForwardSortationAreas")
view(toronto_forward_sortation_areas)

@@

## Summary

As demand for EMS continues to be a challenge, identifying the areas where most of the demand comes from can help better inform strategies to better manage it. Thus, this analysis investigates...

To this, the given analysis finds that medical emergencies and vehicle accidents are notable drivers of EMS demand within Toronto.

@@

## Limitations

CONTENT

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-analysisdata
#| tbl-cap: "Paramedic Services Incident Data Summary Statistics"

@@

summary_data <- summary(data[,2:8])
#view(summary_data)

summary_data |>
    kable(
    col.names = c("Medical Emergencies", "Vehicle Accidents", "Emergency Transfers",
                  "Fires", "Airport Standbys", "Unclassified", "Total Calls"),
    digits = 1,
    booktabs = TRUE,
    linesep = "",
    align = c("l", "l", "l", "l", "l", "l"),
    format.args = list(big.mark = ",")
  )

@@

@tbl-analysisdata presents the summary statistics for each recorded incident type for the years 2017 to 2022.

@@

The shapefile that contains the boundaries for Toronto's FSAs also contained the FSAs for every province and territory of Canada. As this analysis only considers the city of Toronto, ArcGIS Pro software [@arcgis] use employed to remove the spatial information of FSA locations outside Toronto. Specifically, the records of FSA regions that were not in Toronto were removed from the shapefile's attribute table (i.e. data table) using ArcGIS Pro. After modifying this data, it was merged with the data presented in @tbl-mapdata and mapped to show the "EMS Demand Driver" for each FSA. The final map is shown in @fig-map.

The ambulance station location data was not modified for this analysis. It was mapped separately and added to @fig-map to create @(FIG NAME FOR SECOND MAP).
